Constraint Programming - Inside Gecode

### 2014-05-08 paper.tex
corrections

	\subsection{Search Algorithms}
	In constraint programming search algorithms define the way the search tree is constructed, starting from the root node. To improve the performance of the search usually at every newly created node a constraint propagation is executed to narrow the search space down. Based on these two steps then either one, the best or no solution is found. What the actual result in the end is depends on the type of search algorithm.

	In \emph{Gecode} it is possible to define custom strategies, but the user is supported at this point with some standard engines which can be used to build custom ones:
	\begin{itemize}
	\item Depth-First Search
	\item Limited Discrepancy Search
	\item Branch-and-Bound Search
	\item Depth-First Search Restart Optimization
	\end{itemize}

	In the following part I will discuss branching strategies, the Depth-first search and one of its variants, the Branch-and-bound search, the Limited-Discrepancy search as candidate for an improved backjumping-strategy, backjumping strategies in general and restart strategies.

	\subsubsection{Branching Strategies}
	If at one node it is not possible to empty one variable domain or to reduce all variable domains to the size one it is necessary to make a decision: How should the next subtree be constructed? The question is answered by the selected branching strategy. No matter which one is selected, they all have in common that they create additional, temporary constraints, which are only applied to the specific subtrees. This step is called \emph{posting} a constraint on a branch.

	The branching strategy depends on the variable domain: For Integer (and Boolean) variables there are three common strategies \cite[p.~87]{handbookCP} which differ in how the domain is distributed to the subtrees.
	\begin{itemize}
	\item Enumeration: The domain for $x \in D = {x_1, x_2, \dots}$ is split up into $|D|$-branches where for every branch $i$ one additional constraint $c:\ x = x_i$ is posted.
	\item Binary choice points: One value $x_i$ of the domain $D$ of the variable $x$ is selected and the domain is split up into two subtrees: On one the constraint $x = x_i$ and on the other one the constraint $x \ne x_i$ is posted. This results into something like a binary tree at this point.
	\item Domain splitting: The domain $D$ for the variable $x$ is split up into two disjoint intervals or sets, for example $x < 2$ and $x \ge 2$.
	\end{itemize}

	As for variables with a Float domain the Enumeration and Binary choice points are not feasible, the last one, domain splitting, is typically used. As termination criteria for a Float it is often necessary to define a certain epsilon $\epsilon \in \mathbb{R}_{>0}$ which describes the interval length, at which the solver states to have found a solution.

	For all branching strategies there is a choice of freedom left, which can be used to implement some heuristics. This additional information is either gained during the search itself of predefined by the user.

	\subsubsection{Depth-First Search}
	The Depth-first search is a good point to start, as it was also historical the first developed one. The idea behind this search algorithm is that every possible combination of variables is examined. Obviously in its basic form it can not be applied to problems with inifinite domains; for problems with such variable domains it is necessary to use for instance Branch-and-reduce. Sometimes the Depth-first search is also referred to as chronological backtrack search.

	The starting point is the root node, from where on the algorithm is called recursively. At every node mainly two steps are executed: First a constraint propagation and then a branching, which divides the search space into two disjoint sub spaces, on which the same algorithm is applied again. The search stops in a leave if either no branching is possible anymore and therefore a valid solution is found, or one or more variable domains are emptied, which indicates that in this leave no possible solution exists.

@@

	\subsubsection{Depth-First Search with Optimization}
	If in the constraint programming an objective is defined which should be either minimized or maximized a modified version of the Depth-first search can be applied. The only difference is that it evalutes the objective function for every found solution concerning the constraints and stores during the search only the best one. To accelerate the search it is possible to add an additional constraint which defines that the new objective must better than the best so far. Through this enhancement even more sub trees are pruned and the search is faster.

@@

	\subsubsection{Branch-and-Bound Search}
	A branch-and-bound search is an algorithm for optimization of an objective function under certain constraints and only discrete possible candidates for a solution. To fulfill the constraints the first step, branch-and-reduce is applied, and the second one, bounding, is used to prune whole areas with non-optimal solutions.

@@

	As the basic concept of the finding solutions is the enumeration of possible candidates it is necessary to seperate the search space in a useful way, which is the task for the branch-and-reduce step. Actually this step is very similar to the concept of branching, it also depends on a certain heuristic to divide the search space into two or even more disjoint parts. This heuristic can either be selected by the user or be an general one, like splitting the biggest variable domain into two equal sized parts. If the basic search space contains a big amount of possible candidate solution one reduction may not be enough and therefore this step is applied recursively to the newly create subproblems. This process can be stopped if the number of solutions is reduced to an amount, where it is possible to enumerate all of them and check if they are optimal. A possible division into subproblems can be seen in \reffig{fig:branchAndReduce}, where for legibility always only one of the subproblems is divided further.

	The next step after branch-and-reduce is bounding, which is able to prune whole sub problems. In the following part I will consider the minimization of an objective function without a loss of generality. For the bounding it is first necessary to initialize a upper bound with a value, for example with $\infty$. This upper bound is updated during the process every time a new optimal solution is found to this new optimum. If a new possible solution is found which is above this upper bound the algorithm can prune it, as only the minimum solution is of interest. The actually pruning of subproblems is based on the relaxed optimum, which is the optimum of the problem without consideration that only discrete values are allowed. If this relaxed optimum for a certain subproblem is greater than the current upper bound it is possible to prune this sub problem, as it can not contain a better solution.

@@

	\subsubsection{Limited Discrepancy Search}
	Limited Discrepancy Search falls into the category of Best-first searches and was first mentioned by Harvey and Ginsberg in \cite{limitedDiscrepancy}. It depends on some heurisitics for the tree, which are determined through branching strategies. The basic concept is that during a chronological backtrack search a wrong turn close to the root is first of all very expensive. Second, heuristics work worse the farther away from the solution they are applied, therefore decisions close to the root node are possible to fail. Consequently, in a chronological backtrack search the worst decisions are the most expensive ones. The Limited Discrepancy Search solves exactly this problem as it suggests to change rather the decisions made close to the root than the ones the search made later. The improvement compared to a chronological backtrack was shown by experimental results and theoretical proven in \cite{limitedDiscrepancy}.

	\subsubsection{Backjumping}
	A specific version of backjumping was already mentioned, the Limited Discrepancy Search, where instead of jumping one node back the algorithm always jumps as far back to the root node as possible. In general this is proven to be better than the chronological backtracking \cite{limitedDiscrepancy}, but for certain problems it may be possible to find better points to jump back. A jump back in the search tree means to detect a so-called nogood decision, which caused the violation of a constraint in the deadend. How these nogoods must be selected and stored to improve the search depends heavily on the specific field and if it is done wrong the backjumping can even be less efficient than chronological backtracking \cite[p.~100]{handbookCP}.

	\subsubsection{Restart Strategies}
	Branching means always to apply certain heuristics and these heuristics may fail or produce bad results for the first execution. If it turns out that the search tree is build in a disadvantageous way a restart strategy may do the trick to improve the search. For such a restart the heuristic is improved, based on the things the search has learned to be a not so good solution during the first or first few attempts to build a search tree.
	For practical application several variants of restart strategies were proposed \cite[p.~113]{handbookCP}

>>>>

	\subsection{Search Algorithms}
	In constraint programming search algorithms define the way the search tree is constructed and traversed, starting from the root node. To improve the performance of the search usually at every newly created node a constraint propagation is executed to narrow the search space down. Based on these two steps then either one, the best or no solution is found. What the actual result in the end is depends on the type of the search algorithm.

	In \emph{Gecode} it is possible to define custom strategies, but the user is supported at this point with some standard algorithms which can be used to build custom ones:
	\begin{itemize}
		\item Depth-first search
		\item Limited Discrepancy search
		\item Branch-and-bound search
		\item Depth-first search restart optimization
	\end{itemize}

	In the following part I will discuss branching strategies, the Depth-first search, the Branch-and-bound search, the Limited-Discrepancy search as candidate for an improved backjumping-strategy, backjumping strategies in general and restart strategies.

	\subsubsection{Branching Strategies}
	If at one node it is not possible to empty one variable domain or to reduce all variable domains to the size of one it is necessary to make a decision: How should the next subtree be constructed? The question is answered by the selected branching strategy. No matter which one is selected, they all have in common that they create additional, temporary constraints, which are only applied to the specific subtrees. This process is called \emph{posting} a constraint on a branch.

	The branching strategy depends on the variable domain: For Integer (and Boolean) variables there are three common strategies \cite[p.~87]{handbookCP} which differ in how the domain is distributed to the subtrees.
	\begin{itemize}
		\item Enumeration: The domain for $x \in D = {x_1, x_2, \dots}$ is split up into $|D|$-branches where for every branch $i$ one additional constraint $c:\ x = x_i$ is posted.
		\item Binary choice points: One value $x_i$ of the domain $D$ of the variable $x$ is selected and the domain is split up into two subtrees: On one the constraint $x = x_i$ and on the other one the constraint $x \ne x_i$ is posted. This results into something similar to a binary tree at this point.
		\item Domain splitting: The domain $D$ for the variable $x$ is split up into two disjoint intervals or sets, for example $x < 2$ and $x \ge 2$.
	\end{itemize}

	As for variables with a Float domain the enumeration and binary choice points are not feasible, the last one, domain splitting, is typically used. As termination criteria for a Float it is often necessary to define a certain epsilon $\epsilon \in \mathbb{R}_{>0}$ which describes the interval length, at which the solver states to have found a solution.

	For branching strategies there is a choice of freedom left, for instance where the interval is split up, and this freedom can be used to implement some heuristics. This additional information is either gained during the search itself or predefined by the user.

	\subsubsection{Depth-First Search}
	The Depth-first search was historical the first one which was developed. The idea behind this search algorithm is that every possible combination of variables is examined. Obviously in its basic form it can not be applied to problems with inifinite domains; for problems with such variable domains it is necessary to use for instance Branch-and-reduce. Sometimes the Depth-first search is also referred to as chronological backtrack search.

	The starting point is the root node, from where on the algorithm is called recursively. At every node mainly two steps are executed: First a constraint propagation to reduce the variable domains and second a branching, which divides the search space into two disjoint sub spaces, on which the same algorithm is applied again. The search stops in a leave if either no branching is possible anymore and therefore a valid solution is found, or one or more variable domains are emptied, which indicates that in this leave no possible solution exists.

@@

	\subsubsection{Depth-First Search with Optimization}
	If in the constraint programming an objective is defined which should be either minimized or maximized, a modified version of the Depth-first search can be applied. The only difference is that it evalutes the objective function for every found solution concerning the constraints and stores during the search only the best one. To accelerate the search it is possible to add an additional constraint which defines that the new objective must better than the best so far. Through this enhancement even more sub trees are pruned and the search is faster.

@@

	\subsubsection{Branch-and-Bound Search}
	A branch-and-bound search is an algorithm for optimization of an objective function under certain constraints and only discrete possible candidates for a solution. To fulfill the constraints the first step, branch-and-reduce is, applied, and the second one, bounding, is used to prune whole areas with non-optimal solutions.

@@

	As the basic concept for finding solutions in Branch-and-bound is an enumeration of all possible candidates, it is necessary to seperate the search space in a useful way, which is the task for the Branch-and-reduce step. Actually this step is very similar to the concept of branching, it also depends on a certain heuristic to divide the search space into two or even more disjoint parts. This heuristic can either be selected by the user or be a general one, like splitting the biggest variable domain into two equal sized parts. If the basic search space contains a big amount of possible candidate solutions one reduction may not be enough and therefore this step is applied recursively to the newly create subproblems. This process can be stopped if the number of solutions is reduced to an amount, where it is possible to enumerate all of them and check if they are optimal. A possible division into subproblems can be seen in \reffig{fig:branchAndBound}, where for legibility Branch-and-reduce was applied only twice: The first execution split the problem into $P_0$ and the second split the other half into $P_1$ and $P_2$.

	The next step after Branch-and-reduce is the Bounding, which is able to prune whole sub problems. In the following part I will consider the minimization of an objective function without a loss of generality. For the Bounding it is first necessary to initialize a upper bound with a value, for example with $\infty$. This upper bound is updated during the process every time a new optimal solution is found to this new optimum. If a new possible solution is found which is above this upper bound the algorithm can prune it, as only the minimum solution is of interest. The actually pruning of subproblems is based on the relaxed optimum, which is the optimum of the problem without consideration that only discrete values are allowed. If this relaxed optimum for a certain subproblem is greater than the current upper bound it is possible to prune this complete sub problem, as it can not contain a better solution.

@@

	\subsubsection{Limited Discrepancy Search}
	Limited Discrepancy search falls into the category of Best-first searches and was first mentioned by Harvey and Ginsberg in \cite{limitedDiscrepancy}. It depends on some heurisitics for the tree, which are determined through branching strategies. The basic concept is that during a chronological backtrack search a wrong turn close to the root is first of all very expensive. Second, heuristics work worse the farther away from the solution they are applied, therefore decisions close to the root node are likely to fail. Consequently, in a chronological backtrack search the worst decisions are the most expensive ones. The Limited Discrepancy search solves exactly this problem as it suggests to change rather the decisions made close to the root than the ones the search made later. The improvement compared to a chronological backtrack was shown by experimental results and theoretically proven in \cite{limitedDiscrepancy}.

	\subsubsection{Backjumping}
	A specific version of backjumping was already mentioned, the Limited Discrepancy search, where instead of jumping one node back the algorithm always jumps as far back to the root node as possible. In general this is proven to be better than the chronological backtracking \cite{limitedDiscrepancy}, but for certain problems it may be possible to find better nodes to jump back. A jump back in the search tree means to detect a so-called nogood decision, which caused the violation of a constraint in the deadend. How these nogoods must be selected and stored to improve the search depends heavily on the specific field and if it is done wrong the backjumping can even be less efficient than chronological backtracking \cite[p.~100]{handbookCP}.

	\subsubsection{Restart Strategies}
	Branching always means to apply certain heuristics and these heuristics may fail or produce bad results for the first execution of the search algorithm. If it turns out that the search tree is build in a disadvantageous way a restart strategy may do the trick to improve the search. For such a restart the heuristic is improved, based on the things the search has learned to be a not so good solution during the first or first few attempts to build a search tree.
	For practical application several variants of restart strategies were proposed \cite[p.~113]{handbookCP}
